A finite-state morphological transducer for Kyrgyz

### 2012-03-13 paper/paper.tex
LREC template is crap

\author {Jonathan North Washington\\
Departments of Linguistics and Central Eurasian Studies\\
Indiana University\\
Bloomington, IN 47405 (USA)\\
\texttt{jonwashi@indiana.edu} \and
Mirlan Ipasov\\
Computer Engineering and Mathematics Department\\
International Ataturk Alatoo University\\
Bishkek (Kyrgyzstan)\\
\texttt{mipasov@gmail.com} \and
Francis M. Tyers\\
Dept.\ Lleng.\ i Sist.\ Informàtics \\  
Universitat d'Alacant\\
E-03071 Alacant (Spain)\\
\texttt{ftyers@dlsi.ua.es} 
 }

@@

\begin{document}
\maketitle{}
\thispagestyle{empty}

\begin{abstract}
Less than 200 words

\end{abstract}

\section{Introduction}
This paper describes the development of a morphological transducer oriented for the task of machine translation for the Kyrgyz language using the free/open-source platform HFST. The transducer was developed under the auspices of the Apertium \cite{forcada2011} project for use in a machine translation system from Turkish to Kyrgyz.

@@

\section{Background}
\subsection{The Kyrgyz language}
Kyrgyz (written ‹кыргыз тили› or ‹\textuighur{قىرعىز تىلى}›, pronounced [qɯɾʁɯz tili]) alternatively written ``Kirghiz'' or ``Kirgiz'') is a Turkic language spoken in Kyrgyzstan, China, Tajikistan, and Uzbekistan.  Its classification within Turkic remains problematic—it appears to alternatively belong to the Kypchak (Northwestern) branch and to the South Siberian (Northeastern) branch.  The Turkic varieties phonetically and phonologically most similar to Kyrgyz are the southern dialects of Altay, though Kyrgyz shows strong parallels to Kazakh that these varieties lack, especially in its Talas dialects.  In southern varieties of Kyrgyz there are also many similarities to Uzbek that other dialects lack.

@@

>>>>

\author{Jonathan North Washington\\
Departments of Linguistics and Central Eurasian Studies\\
Indiana University\\
Bloomington, IN 47405 (USA)\\
\texttt{jonwashi@indiana.edu} \and
Mirlan Ipasov\\
Computer Engineering and Mathematics Department\\
International Ataturk Alatoo University\\
Bishkek (Kyrgyzstan)\\
\texttt{mipasov@gmail.com} \and
Francis M. Tyers\\
Dept.\ Lleng.\ i Sist.\ Informàtics \\  
Universitat d'Alacant\\
E-03071 Alacant (Spain)\\
\texttt{ftyers@dlsi.ua.es} 
 }

\name{Jonathan North Washington, Mirlan Ipasov, Francis M.\ Tyers}

\address{Departments of Linguistics and Central Eurasian Studies\\
Indiana University\\
Bloomington, IN 47405 (USA)\\
\texttt{jonwashi@indiana.edu} \and
Computer Engineering and Mathematics Department\\
International Ataturk Alatoo University\\
Bishkek (Kyrgyzstan)\\
\texttt{mipasov@gmail.com} \and
Dept.\ Lleng.\ i Sist.\ Informàtics \\  
Universitat d'Alacant\\
E-03071 Alacant (Spain)\\
\texttt{ftyers@dlsi.ua.es}
}

@@

%FIXME: this should be closer to the 200 word limit, and less like the intro pargraph
\abstract{This paper describes the development of a morphological transducer oriented for the task of machine translation for the Kyrgyz language using the free/open-source platform HFST. The transducer was developed under the auspices of the Apertium \citep{forcada2011} project for use in a machine translation system from Turkish to Kyrgyz.}

%FIXME:
%\thanks{GSoC, Tolgonay, ..?}

\begin{document}
\maketitleabstract{}
\thispagestyle{empty}

\section{Introduction}
This paper describes the development of a morphological transducer oriented for the task of machine translation for the Kyrgyz language using the free/open-source platform HFST. The transducer was developed under the auspices of the Apertium \citep{forcada2011} project for use in a machine translation system from Turkish to Kyrgyz.

@@

\section{Background}
\subsection{The Kyrgyz language}
Kyrgyz (written ‹кыргыз тили› or ‹\textuighur{قىرعىز تىلى}›, pronounced [qɯɾʁɯz tilí]) alternatively written ``Kirghiz'' or ``Kirgiz'') is a Turkic language spoken in Kyrgyzstan, China, Tajikistan, and Uzbekistan.  Its classification within Turkic remains problematic—it appears to alternatively belong to the Kypchak (Northwestern) branch and to the South Siberian (Northeastern) branch.  The Turkic varieties phonetically and phonologically most similar to Kyrgyz are the southern dialects of Altay, though Kyrgyz shows strong parallels to Kazakh that these varieties lack, especially in its Talas dialects.  In southern varieties of Kyrgyz there are also many similarities to Uzbek that other dialects lack.

@@

%FIXME: X and Y
Current morphological analysers of European languages are based on the orthography of the words, even in cases such as X and Y where this may make it more difficult to write morphophonological rules.  This has the advantage that in order to use the morphological analyser to analyse text (as opposed to using it as a tool to study phonology), no up/down conversion between the orthography and the transcription used in the analyser is necessary, avoiding possibilities of misconversion.

In the case of Kyrgyz, dealing with the orthographical forms directly further simplifies some aspects of the morphophonology, since Kyrgyz orthography reflects a somewhat simplified version of the phonology: it ignores processes that the orthographies of many other Turkic do not, such as the phonemic distinction between velar and uvular consonants as well as sandhi voicing effects.
